Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-28 Manuscript.Rmd
Update Manuscript.Rmd

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We inferred different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring. While oligotrophic species were more likely to colonize compared to random expectation, we found only weak shifts in plant community structure. Particularly, the average nutrient value of plant communities remained stable over time. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking average community composition.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low Nitrogen availability [@Vitousek1997], the addition of Nitrogen is likely to alter these communities. Indeed, Nitrogen deposition together with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased Nitrogen deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load [@Slootweeg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen deposition [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

Recovery of extant plant communities from high Nitrogen deposition would imply that the state of communities measured at different time points is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness is an easy to measure and communicate biodiversity endpoint and it is often negatively correlated with Nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference has often been expressed using indicator values assigned to each plant species [@Ellenberg1992]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus a useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70% of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are also of importance agronomically [@Leiber2006]. In mountain hay meadows, the spatial variation of species richness in vascular plants is negative correlated with Nitrogen deposition [@Roth2013], suggesting that mountain grasslands suffered from increased Nitrogen deposition. However, between 1990 and 2010 NO$_x$ emissions in Switzerland decreased by 46% and NH$_3$ emissions by 14% [@MaasGrennfelt2016] and we expect plant communities to recover.

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at the landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the result of recent climate warming interacting with airborne nitrogen deposition [@Steinbauer2018]. Furthermore, traditional management regimes are currently changing, which also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, Nitrogen deposition or land-use intensity change over the last 15 years, (2) was species turnover correlated with the average temperature, precipitation, Nitrogen deposition or inclination (we expect steep areas to be less intensively managed), and (3) did species that newly colonized or disappeared (local extinction) from sites differ from random expectation according to their indicator values for temperature, humidity, nutrients or light.

@@

After the sampling of the plant data the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria resulted in sample of `r nrow(sites)` sites.

For each survey (a survey consists of the two visits per season) we calculated the following biodiversity endpoints [@Rowe2017]: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for (2) temperature, (3) humidity, (4) nutrients and (5) light. The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. Additionally, to these five biodiversity endpoints that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points to describe the community change between two time points.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First, we expected communities to response to climate warming and the response might be different along the temperature gradient [@Roth2014]. To describe the temperature gradient, we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see also @RihmAchermann2016 and @Roth2017. Finally, we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

@@

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during first survey and recorded during second or not recorded during second and recorded during third). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. Then we calculated the differences of the CM minus the average of the random-CMs to obtain a standardized measure (standardized-CM) of how different the disappearing species were from random expectation. A standardized-CM < 0 would suggest that the indicator values of colonizing species were lower than the indicator values of average species that were occurring at this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients and light (see Appendix A). We then inferred whether the standardized-CM is changing along the corresponding gradient using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic-LMM to infer whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this changing depended on the species indicator value for nutrients. To analyse the colonization probability, we selected all species that were not observed during the first survey and asked if they were observed (then Occ = 1) or not observed (then Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey if they were observed or not observed (then Occ = 0) during the third survey. The variable 'occ' was then used as dependent variable in the logistic-LMM. As predictor variables the model contained the Nitrogen deposition of the site, the indicator value for nutrients of the species and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic LMM was also used to infer local survival probability. In that case, however, we selected all species that were recorded during the first or second survey and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

To estimate the effect of Nitrogen deposition on total species richness at a given time point, we described the plant species richness at the sites using a generalized linear model with Poisson distribution and the logarithm as link function. As predictors we used the four environmental gradients as described above. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

>>>>

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in N-deposition. We considered different drivers of plant community change (i.e. N-deposition, climate warming, land-use change) in Swiss mountain hay meadows using data from the Swiss biodiversity monitoring. While oligotrophic plant species were more likely to colonize compared to random expectation, we found only weak shifts in plant community structure. Particularly, the average nutrient value of plant communities remained stable over time. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking average community composition.

@@

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via input from the atmosphere to the biosphere. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low Nitrogen availability [@Vitousek1997], the addition of Nitrogen is likely to alter these communities. Indeed, Nitrogen deposition together with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased Nitrogen deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully diminished emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that suffered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load [@Slootweeg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear subsequent to reduced N-deposition [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in N-deposition rates will lead to the recovery of extant plant communities.

Recovery of extant plant communities from high N- deposition would imply that the state of communities measured at different points in time is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness is an easy to measure and communicate biodiversity endpoint and it is often negatively correlated with Nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more useful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference has often been expressed using indicator values assigned to each plant species [@Ellenberg1992]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus a useful approach to understanding biodiversity change is through estimates of species turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70% of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are also of agronomical importance [@Leiber2006]. In mountain hay meadows, the spatial variation of species richness in vascular plants is negative correlated with Nitrogen deposition [@Roth2013], suggesting that mountain grasslands suffered from increased Nitrogen deposition. However, between 1990 and 2010 NO$_x$ emissions in Switzerland decreased by 46% and NH$_3$ emissions by 14% [@MaasGrennfelt2016] and we would expect plant communities to recover.

In addition to N-deposition, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warming rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at the landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the result of recent climate warming interacting with airborne nitrogen deposition [@Steinbauer2018]. Furthermore, traditional management regimes are currently changing, which also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. Management regimes of accessible mountainous areas are frequently intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, Nitrogen deposition or land-use intensity change over the last 15 years, (2) was species turnover correlated with the average temperature, precipitation, Nitrogen deposition or inclination (we expect steep areas to be less intensively managed), and (3) did species that newly colonized or disappeared (local extinction) from sites differ from random expectation according to their indicator values for temperature, humidity, nutrients or light.

@@

After the sampling of the plant data the botanists also assigned a habitat type to each sampling site according to the classification system developed for Switzerland [@Delarze2008]. We matched the habitat types of the Swiss classification system with the categories from the EUNIS system [level-3 classification; @Davies2004] and selected all sampling sites in mountain hay meadows (EUNIS E2.3). We analysed the data from 2003 to 2017. During that study period each sampling site was surveyed once per five-year period: the first period lasted from 2003 to 2007, the second from 2008 to 2012 and the third from 2013 to 2017. These selection criteria yielded a sample of `r nrow(sites)` sites.

For each survey (a survey consists of two  visits per season) we calculated the following biodiversity endpoints [@Rowe2017]: (1) The number of recorded species (species richness), the community mean of indicator values of recorded species for (2) temperature, (3) humidity, (4) nutrients and (5) light. The indicator values were obtained from the recalibrated indicator values for the Swiss Flora [@Landolt2010]. In addittion to these five biodiversity endpoints that describe the state of plant communities for each site at a given time point, we also estimated the temporal turnover (i.e. species exchange ratio sensu @Hillebrand2018) as the proportion of species that differ between two time points to describe the community change between two points in time.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First, we expected communities to response to climate warming with this response being dependent on the position of the site along the temperature gradient [@Roth2014]. To describe the temperature gradient, we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key driver for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see also @RihmAchermann2016 and @Roth2017. Finally, we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

@@

To infer whether species that colonized or disappeared from a site had particular indicator values that differed from the other species at that site, we produced for each site a list with all species that were recorded during the three surveys (total community). We then calculated the community mean (CM) of the indicator value for all species that colonized the site during the three surveys (i.e. not recorded during first survey and recorded during second or not recorded during second and recorded during third). We then randomly selected the same number of species from the total community and also calculated the community mean of the value for these species (random-CM). We repeated the random selection of species 1000 times. Then we calculated the differences of the CM minus the average of the random-CMs to obtain a standardized measure (standardized-CM) of how different the disappearing species were from random expectation. For example, a standardized-CM < 0 would suggest that the indicator values of colonizing species were lower than might be expected from random colonisation from the species-pool for this site. We applied this method for both colonizing and disappearing species and for the indicator values for temperature, humidity, nutrients and light (see Appendix A). We then tested whether the standardized-CM is changing along the corresponding gradient (inferred from independet datasets, see Environmental gradients section) using linear models. Model parameters were estimated in a Bayesian framework using the R-Package *rstanarm* [@Stan2016, @Muth2018].

We used logistic-LMM to analyse whether the colonization probability or local survival probability was changing along the Nitrogen deposition gradient and whether this change depended on the species indicator value for nutrients. To analyse the colonization probability of species, we selected all species that were not observed during the first survey. For each of these species we asked if they were observed (then Occ = 1) or not observed (then Occ = 0) during the second survey. The same was also done for all species that were not observed during the second survey if they were observed or not observed (then Occ = 0) during the third survey. The variable 'occ' was then used as dependent variable in the logistic-LMM. As predictor variables the model contained the Nitrogen deposition of the site, the indicator value for nutrients of the species and the interaction of these two variables. Additionally, species-ID and site-ID were included as random effects. The same logistic LMM was also used to investigate local survival probabilities  . In that case, however, we selected all species that were recorded during the first or second survey and the variable 'occ' then indicated whether or not the species was also observed during the next survey. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].

To estimate the effect of Nitrogen deposition on total species richness at a given point in time, we described the plant species richness at the sites using a generalized linear model with Poisson distribution and the logarithm as link function. As predictors we used the four environmental gradients as described above. Model parameters were estimated using an approximate Bayesian approach using the R-Package *arm* [@arm2018].
